Death by Dropdown?<br>Engineer Insightful Shiny Apps<br>with Behavioral Science

### 2025-09-13 conference-materials/2025_posit-conf/draft.qmd
Final revamp before finishing touches
- Deleted `summary-report.R`, `telemetry_heatmap.R`, and associated `dashboard_telemetry.sqlite` data file, as telemetry visualization is now pre-rendered.
- Removed `bidux-outline.md`, with content now embedded in `draft.qmd`.
- Updated slide assets and HTML to reflect finalized UI screenshots and static plots.
- Simplified workflow by removing runtime dependencies and focusing on pre-built presentation content.

::: {.column width="50%" .center}
### Welcome! You're among friends 😊 {style="color: var(--coral-light) !important;"}

@@

## {background-image="img/ui-before.png" background-size="contain" background-color="#2c3e50"}

@@

- 18 filters across 6 tabs
- Every possible data view
- Zero user adoption

@@

::: {.notes}
- 45 seconds
- I worked with a client who had this 'comprehensive' dashboard
- Blinded here for confidentiality...but incredibly close to the actual design
- 18 filters, 6 tabs, every possible view, nearly zero adoption
- Why this matters: This represents thousands of developer hours and client budget wasted on technically perfect but psychologically broken software
- The client was frustrated bc they had spent months building this thing, but no one used it
- As developers, we can relate...we've all built features nobody uses
- But here's why this matters more than frustration:
    1. Technical debt from unused complexity compounds over time
    2. User abandonment patterns replicate across similar projects
    3. Without data-driven insights, we keep making the same UX mistakes
- [CLICK] So we ran telemetry for 30 days...
- This data changed everything about how we approached the dashboard design
:::

@@

library(ggplot2)
library(dplyr)

# Create telemetry data with journey visualization
telemetry_data <- data.frame(
  metric = factor(
    c(
      "Opened dashboard",
      "Clicked any filter",
      "Used >3 filters",
      "Found what they needed",
      "Came back next day"
    ),
    levels = c(
      "Opened dashboard",
      "Clicked any filter",
      "Used >3 filters",
      "Found what they needed",
      "Came back next day"
    )
  ),
  percentage = c(100, 35, 18, 8, 12),
  dropoff = c(0, 65, 17, 10, -4)
)

ggplot(telemetry_data, aes(x = metric, y = percentage)) +
  geom_col(
    fill = c("#27ae60", "#e74c3c", "#c0392b", "#a93226", "#e74c3c"),
    width = 0.7
  ) +
  geom_text(
    aes(label = paste0(percentage, "%")),
    vjust = -0.5,
    size = 10,
    fontface = "bold",
    # color = "white"
  ) +
  geom_segment(
    aes(
      x = as.numeric(metric) - 0.35,
      xend = as.numeric(metric) + 0.35,
      y = percentage,
      yend = percentage
    ),
    # color = "white",
    size = 1,
    alpha = 0.3
  ) +
  labs(
    subtitle = "1,847 sessions tracked over 30 days",
    x = "",
    y = "Percentage of Users"
  ) +
  theme_minimal(base_size = 20) +
  theme(
    # plot.background = element_rect(fill = "#e74c3c", color = NA),
    # panel.background = element_rect(fill = "#e74c3c", color = NA),
    plot.title = element_text(size = 36, face = "bold"),
    plot.subtitle = element_text(size = 24),
    axis.text.x = element_text(
      size = 16,
      angle = 25,
      hjust = 1
    ),
    axis.text.y = element_text(size = 18),
    # axis.title = element_text(color = "white"),
    panel.grid = element_blank()
  ) +
  ylim(0, 110)
```

::: {.fragment .text-xl style="text-align: center"}
**92% of users failed to find what they needed**
:::

::: {.notes}
- 45 seconds
- Look at this journey of despair
- 100% open, only 35% click anything
- Only 8% succeed in finding what they needed
- And we didn't collected data on this..but I would hypothesize that many of those folks were a bit more data savvy and just knew where to look
- [CLICK] Overall...this was essentially 92% failure rate on a product that cost the client thousands of dollars to build
- This is behavioral science **screaming** at us that something is wrong here
:::

@@

library(ggplot2)

# Enhanced heatmap
filters <- data.frame(
  filter = factor(
    paste("Filter", 1:18),
    levels = rev(paste("Filter", 1:18))
  ),
  usage = c(82, 78, 71, 12, 8, 6, 5, 3, 3, 2, 2, 1, 1, 1, 0, 0, 0, 0)
)

ggplot(filters, aes(x = 1, y = filter, fill = usage)) +
  geom_tile(color = "white", size = 1) +
  geom_text(
    aes(label = paste0(usage, "%")),
    size = 5, fontface = "bold",
    color = ifelse(filters$usage > 40, "white", "black")
  ) +
  scale_fill_gradient2(
    low = "#fee5d9", mid = "#fc9272", high = "#a50f15",
    midpoint = 40, guide = "none"
  ) +
  labs(
    title = "Filter Usage Reality",
    subtitle = "14 of 18 filters basically ignored",
    x = "", y = ""
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    plot.title = element_text(size = 20, face = "bold"),
    plot.subtitle = element_text(size = 16, color = "#666")
  )
```
:::

@@

## Complete Workflow {background-color="#2c3e50"}

``` {r}
#| echo: false
invisible({
  capture.output({
    issues <- bid_telemetry("dashboard_telemetry.sqlite")
  })
})
```

```{r}
#| eval: false
#| code-line-numbers: "1-2|4-6|7|8-10|11"

@@

## Complete Workflow {background-color="#2c3e50"}

```{r}
#| code-line-numbers: "1|2-10|11-14|15-17|18|19"
#| output-location: fragment

# Without telemetry data
report <- bid_interpret(
    central_question = "Which markets are driving performance?",
    data_story = list(
      hook = "Q4 revenue hit record high, but satisfaction dipped",
      context = "After aggressive marketing across all regions",
      tension = "West region satisfaction fell 10 points",
      resolution = "Focus retention efforts on underperforming regions"
    )
  ) |>
  bid_notice(
    problem = "Users are struggling to find key insights",
    evidence = "Users report not knowing where to start"
  ) |>
  bid_anticipate() |>
  bid_structure() |>
  bid_validate() |>
  bid_report(format = "text", include_diagrams = FALSE)

print(report)
```

@@

![](img/ui-before.png)

::: {.notes}
- 45 seconds
- "Before: 18 filters, 8% success"
- "After: 3 smart filters, 85% success"
- [CLICK] "Same data, 10x better outcomes"
:::

::: {.text-box-overlay-dark .fragment}
| Metric          | Before     | After              |
|---------------- |:---------: |:------------------:|
| Filters         | 18         | 3 (+15 advanced)   |
| Success Rate    | 8%         | 85%                |
| Time to Insight | 258s       | 45s                |
| Return Rate     | 12%        | 67%                |

::: {.fragment .text-xl style="margin-top: 30px;"}
**Same data. Same goals. 10x better outcomes.**
:::
:::
::::

::: {.notes}
- 45 seconds
- "Before: 18 filters, 8% success"
- "After: 3 smart filters, 85% success"
- [CLICK] "Same data, 10x better outcomes"
:::

library(ggplot2)
library(tidyr)

# Enhanced metrics visualization
metrics <- data.frame(
  Metric = factor(
    c("Time to Insight", "Task Completion", "User Satisfaction", "Return Rate"),
    levels = c(
      "Time to Insight",
      "Task Completion",
      "User Satisfaction",
      "Return Rate"
    )
  ),
  Before = c(258, 35, 32, 12),
  After = c(45, 85, 78, 67),
  Industry = c("All", "All", "All", "All")
)

metrics_long <- metrics |>
  pivot_longer(cols = c(Before, After), names_to = "State", values_to = "Value")

# Calculate improvements
metrics_long$Improvement <- ifelse(
  metrics_long$State == "After",
  round((metrics_long$Value / metrics$Before - 1) * 100),
  NA
)

ggplot(metrics_long, aes(x = Metric, y = Value, fill = State)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_text(
    aes(
      label = ifelse(
        Metric == "Time to Insight",
        paste0(Value, "s"),
        paste0(Value, "%")
      )
    ),
    position = position_dodge(width = 0.7),
    vjust = -0.5, size = 8, fontface = "bold") +
  geom_text(
    data = subset(metrics_long, State == "After" & !is.na(Improvement)),
    aes(label = paste0("+", abs(Improvement), "%"), y = Value + 10),
    position = position_dodge(width = 0.7),
    vjust = -0.5, size = 6, color = "#27ae60", fontface = "italic") +
  scale_fill_manual(values = c("Before" = "#e74c3c", "After" = "#27ae60")) +
  labs(
    title = "BID Framework Impact: 12 Enterprise Deployments",
    subtitle = "Average improvements across healthcare, finance, and retail sectors",
    y = ""
  ) +
  theme_minimal(base_size = 20) +
  theme(
    legend.position = "top",
    legend.text = element_text(size = 24),
    legend.title = element_blank(),
    plot.title = element_text(size = 32, face = "bold"),
    plot.subtitle = element_text(size = 24),
    axis.text = element_text(size = 20),
    axis.title = element_blank(),
    panel.grid.major.x = element_blank()
  ) +
  ylim(0, max(metrics_long$Value) * 1.2)
```

@@

::: {.column width="35%" .fragment}
### Resources:
- 🔬 **BID Framework**: [github.com/jrwinget/bid-framework](https://github.com/jrwinget/bid-framework)
- 📚 **{bidux} Docs**: [github.com/jrwinget/bidux](https://github.com/jrwinget/bidux)
- 💬 **Community**: [github.com/jrwinget/bidux/discussions](https://github.com/jrwinget/bidux/discussions)
- 🤝 **Let's Connect!**
  - <svg xmlns="http://www.w3.org/2000/svg" width="16" height="16" fill="currentColor" class="bi bi-bluesky" viewBox="0 0 16 16">
  <path d="M3.468 1.948C5.303 3.325 7.276 6.118 8 7.616c.725-1.498 2.698-4.29 4.532-5.668C13.855.955 16 .186 16 2.632c0 .489-.28 4.105-.444 4.692-.572 2.04-2.653 2.561-4.504 2.246 3.236.551 4.06 2.375 2.281 4.2-3.376 3.464-4.852-.87-5.23-1.98-.07-.204-.103-.3-.103-.218 0-.081-.033.014-.102.218-.379 1.11-1.855 5.444-5.231 1.98-1.778-1.825-.955-3.65 2.28-4.2-1.85.315-3.932-.205-4.503-2.246C.28 6.737 0 3.12 0 2.632 0 .186 2.145.955 3.468 1.948"/></svg> [@jrwinget](https://bsky.app/profile/jrwinget.bsky.social)
  - <svg xmlns="http://www.w3.org/2000/svg" width="16" height="16" fill="currentColor" class="bi bi-linkedin" viewBox="0 0 16 16">
  <path d="M0 1.146C0 .513.526 0 1.175 0h13.65C15.474 0 16 .513 16 1.146v13.708c0 .633-.526 1.146-1.175 1.146H1.175C.526 16 0 15.487 0 14.854zm4.943 12.248V6.169H2.542v7.225zm-1.2-8.212c.837 0 1.358-.554 1.358-1.248-.015-.709-.52-1.248-1.342-1.248S2.4 3.226 2.4 3.934c0 .694.521 1.248 1.327 1.248zm4.908 8.212V9.359c0-.216.016-.432.08-.586.173-.431.568-.878 1.232-.878.869 0 1.216.662 1.216 1.634v3.865h2.401V9.25c0-2.22-1.184-3.252-2.764-3.252-1.274 0-1.845.7-2.165 1.193v.025h-.016l.016-.025V6.169h-2.4c.03.678 0 7.225 0 7.225z"/></svg> [@jrwinget](https://www.linkedin.com/in/jrwinget/)
  - 📧 [contact@jrwinget.com](mailto:contact@jrwinget.com)
:::
::::

>>>>

::: {.column width="50%" .center .fragment}
### Welcome! You're among friends 😊 {style="color: var(--coral-light) !important;"}

@@

## {background-image="img/ui-before1.png" background-size="contain" background-color="#2c3e50"}

@@

- 18 filters across 6 tabs
- Every possible data view
- Users' curiosity quickly vanished
    - And so did their visits

@@

::: {.notes}
- 45 seconds
- I worked with a client who had this 'comprehensive' dashboard
- Blinded here for confidentiality...but incredibly close to the actual design
- 18 filters, 6 tabs, every possible view, nearly zero adoption
- Why this matters: This represents thousands of developer hours and client budget wasted on technically perfect but psychologically broken software
- The client was frustrated bc they had spent months building this thing, but no one used it
- Users were initially curious, but quickly overwhelmed...and eventually, they just vanished
- As developers, we can relate...we've all built features nobody uses
- But here's why this matters more than frustration:
    1. Technical debt from unused complexity compounds over time
    2. User abandonment patterns replicate across similar projects
    3. Without data-driven insights, we keep making the same UX mistakes
- [CLICK] So we ran telemetry for 30 days...
- This data changed everything about how we approached the dashboard design
:::

@@

library(ggplot2)
library(dplyr)
library(tidyr)

journey <- tibble(
  stage = c(
    "Visited app URL",
    "Opened dashboard",
    "Applied ≥1 filter",
    "Returning within 1 week"
  ),
  pct   = c(95, 88, 52, 20)
)

journey |>
  mutate(stage = reorder(stage, -pct)) |>
  ggplot(aes(x = stage, y = pct)) +
  geom_col(fill = "#5FBDB0") +
  geom_text(
    aes(label = paste0(pct, "%")),
    vjust = -0.5, size = 10
  ) +
  scale_y_continuous(limits = c(0,100), expand = c(0,0)) +
  labs(
    subtitle = "1,287 sessions tracked over 30 days",
    y = "Users (%)", x = NULL
  ) +
  theme_minimal(base_size = 20) +
  theme(
    plot.title = element_text(size = 36, face = "bold"),
    plot.subtitle = element_text(size = 24),
    axis.text.x = element_text(
      size = 16,
      angle = 25,
      hjust = 1
    ),
    axis.text.y = element_text(size = 18),
    panel.grid = element_blank()
  ) +
  ylim(0, 110)
```

::: {.notes}
- 45 seconds
- Look at this journey of despair
- [update notes here]
- This is behavioral science **screaming** at us that something is wrong here
:::

@@

filters <- tibble::tibble(
  filter = factor(
    paste("Filter", 1:18),
    levels = rev(paste("Filter", 1:18))
  ),
  pct = c(
    33.8, 21.4, 13.6, 7.8, 4.9, 3.9, 2.9, 2.4, 1.8,
    1.4, 1.2, 1.0, 0.9, 0.8, 0.7, 0.6, 0.5, 0.4
  )
)

ggplot(filters, aes(x = 1, y = filter, fill = pct)) +
  geom_tile(color = "white", size = 1) +
  geom_text(
    aes(label = paste0(pct, "%")),
    size = 5, fontface = "bold",
    color = ifelse(filters$pct > 40, "white", "black")
  ) +
  scale_fill_gradient2(
    low = "#FCE788", mid = "#FCA311", high = "#D7263D",
    midpoint = 40, guide = "none"
  ) +
  labs(
    title = "Actual Filter Usage",
    subtitle = "14 of 18 filters basically ignored",
    x = "",
    y = ""
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    plot.title = element_text(size = 20, face = "bold"),
    plot.subtitle = element_text(size = 16, color = "#666")
  )
```
:::

@@

![](img/ui-before2.png)

::: {.notes}
- 45 seconds
- [update notes here]
- "Δ (pp)"" is absolute percentage-point change
- Relative change is After/Before - 1, then rounded
- [CLICK] More users. Deeper engagement. Stronger retention.
:::

::: {.text-box-overlay-dark .fragment}
| Metric                  | Before (%) | After (%) |   Δ    | Relative change |
| ----------------------- | :--------: | :-------: | :----: | :-------------- |
| Visited app URL         |     95     |     97    |   +2   | +2%             |
| Opened dashboard        |     88     |     95    |   +7   | +8%             |
| Applied ≥1 filter       |     52     |     72    |   +20  | +38%            |
| Returning within 1 week |     20     |     35    |   +15  | +75%            |

::: {.fragment .text-xl style="margin-top: 30px;"}
**More users. Deeper engagement. Stronger retention.**
:::
:::
::::

::: {.notes}
- 45 seconds
- [update notes here]
- "Δ (pp)"" is absolute percentage-point change
- Relative change is After/Before - 1, then rounded
- [CLICK] More users. Deeper engagement. Stronger retention.
:::

n_deploy <- 12

metrics_long <- tibble(
  Metric = factor(
    c("Time to Insight", "Task Completion", "User Satisfaction", "Return Rate"),
    levels = c("Time to Insight", "Task Completion", "User Satisfaction", "Return Rate")
  ),
  Before = c(180, 48, 44, 22),
  After  = c(115, 62, 56, 30)
) |>
  pivot_longer(cols = c(Before, After),
               names_to = "State", values_to = "Value") |>
  mutate(State = factor(State, levels = c("Before", "After"))) |>
  group_by(Metric) |>
  mutate(
    BeforeValue = Value[State == "Before"][1],
    DiffPP      = ifelse(State == "After", Value - BeforeValue, NA_real_),
    MaxValue    = max(Value)
  ) |>
  ungroup()

pal <- c("Before" = "#4C78A8", "After" = "#72B7B2")

ggplot(metrics_long, aes(x = Metric, y = Value, fill = State)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.7) +
  geom_text(
    aes(
      label = ifelse(
        Metric == "Time to Insight",
        paste0(Value, "s"),
        paste0(Value, "%")
      )
    ),
    position = position_dodge(width = 0.7),
    vjust = -0.5,
    size = 6,
    fontface = "bold"
  ) +
  geom_text(
    data = metrics_long |>
      filter(State == "After" & !is.na(DiffPP)) |>
      distinct(Metric, .keep_all = TRUE),
    aes(
      y = MaxValue + 20,
      label = ifelse(
        Metric == "Time to Insight",
        paste0(round(DiffPP), "s"),
        paste0("+", round(DiffPP), "%")
      )
    ),
    color = "#FF4081",
    size = 6,
    fontface = "bold"
  ) +
  scale_fill_manual(values = pal) +
  labs(
    title = paste("BID Framework Impact:", n_deploy, "Enterprise Deployments"),
    subtitle = "Average across healthcare, finance, and customer goods sectors",
    y = NULL,
    x = NULL
  ) +
  theme_minimal(base_size = 18) +
  theme(
    legend.position = "top",
    legend.title = element_blank(),
    plot.title = element_text(size = 24, face = "bold"),
    plot.subtitle = element_text(size = 14),
    axis.text = element_text(size = 14),
    panel.grid.major.x = element_blank()
  ) +
  coord_cartesian(
    ylim = c(0, max(metrics_long$MaxValue) + 20)
  )
```

## Simplified UX Workflows {background-color="#2c3e50"}

```{r}
#| eval: false
#| code-line-numbers: "1-2|4-5|6|7|8-10|11"

@@

## Simplified UX Workflows {background-color="#2c3e50"}

```{r}
#| eval: false
#| code-line-numbers: "1|2-10|11-14|15-17|18"

# Without telemetry data
report <- bid_interpret(
    central_question = "Which markets are driving performance?",
    data_story = list(
      hook = "Q4 revenue hit record high, but satisfaction dipped",
      context = "After aggressive marketing across all regions",
      tension = "West region satisfaction fell 10 points",
      resolution = "Focus retention efforts on underperforming regions"
    )
  ) |>
  bid_notice(
    problem = "Users are struggling to find key insights",
    evidence = "Users report not knowing where to start"
  ) |>
  bid_anticipate() |>
  bid_structure() |>
  bid_validate() |>
  bid_report(format = "text", include_diagrams = FALSE)
```

@@

@@

::: {.column width="35%" .fragment}
### Resources:
- 🔬 **BID Framework**: [github.com/jrwinget/bid-framework](https://github.com/jrwinget/bid-framework)
- 📚 **{bidux} Docs**: [github.com/jrwinget/bidux](https://github.com/jrwinget/bidux)
- 💬 **Community**: [github.com/jrwinget/bidux/discussions](https://github.com/jrwinget/bidux/discussions)
- 🤝 **Let's Connect!**
  - <svg xmlns="http://www.w3.org/2000/svg" width="16" height="16" fill="currentColor" class="bi bi-bluesky" viewBox="0 0 16 16">
  <path d="M3.468 1.948C5.303 3.325 7.276 6.118 8 7.616c.725-1.498 2.698-4.29 4.532-5.668C13.855.955 16 .186 16 2.632c0 .489-.28 4.105-.444 4.692-.572 2.04-2.653 2.561-4.504 2.246 3.236.551 4.06 2.375 2.281 4.2-3.376 3.464-4.852-.87-5.23-1.98-.07-.204-.103-.3-.103-.218 0-.081-.033.014-.102.218-.379 1.11-1.855 5.444-5.231 1.98-1.778-1.825-.955-3.65 2.28-4.2-1.85.315-3.932-.205-4.503-2.246C.28 6.737 0 3.12 0 2.632 0 .186 2.145.955 3.468 1.948"/></svg> [@jrwinget](https://bsky.app/profile/jrwinget.bsky.social)
  - <svg xmlns="http://www.w3.org/2000/svg" width="16" height="16" fill="currentColor" class="bi bi-linkedin" viewBox="0 0 16 16">
  <path d="M0 1.146C0 .513.526 0 1.175 0h13.65C15.474 0 16 .513 16 1.146v13.708c0 .633-.526 1.146-1.175 1.146H1.175C.526 16 0 15.487 0 14.854zm4.943 12.248V6.169H2.542v7.225zm-1.2-8.212c.837 0 1.358-.554 1.358-1.248-.015-.709-.52-1.248-1.342-1.248S2.4 3.226 2.4 3.934c0 .694.521 1.248 1.327 1.248zm4.908 8.212V9.359c0-.216.016-.432.08-.586.173-.431.568-.878 1.232-.878.869 0 1.216.662 1.216 1.634v3.865h2.401V9.25c0-2.22-1.184-3.252-2.764-3.252-1.274 0-1.845.7-2.165 1.193v.025h-.016l.016-.025V6.169h-2.4c.03.678 0 7.225 0 7.225z"/></svg> [@jrwinget](https://www.linkedin.com/in/jrwinget/)
  - 🌐 [jrwinget.com](https://www.jrwinget.com)
  - 📧 [contact@jrwinget.com](mailto:contact@jrwinget.com)
:::
::::
